

\chapter{Array Lists}
\label{chap-arraylist}
%TODO Find the Ur List and Ur dynamic array
The first data structure we will be studying is the list.
The list is by far the most relatable data structure, as humans deal with lists on a regular basis.

\section{What is a List?}


When you get right down to it, lists are defined by order.
We don't have to take advantage of this order, but its there.
Populated lists have a first item and they have a last item.

Take a look at this quest below from a hypothetical fantasy game:

\subsubsection*{Quest: Slay the Dragon of Doom}
\begin{itemize}
	\item Get Sword of Dragonslaying
	\item Locate the map to Dragon Lair of Doom
	\item Travel to the Dragon Lair of Doom 
	\item Slay the Dragon of Doom
	\item Return to the Castle
\end{itemize}

Here, the order is implied by the contents of the list - you can't beat the dragon without the macguffin and you certainly can't fight it without being able to find it.
Generally speaking, going up against a dragon without any preparation is foolhardy in the extreme, but I digress.

Thus, you must get the special sword\footnote{What if its possible to get the map before the sword? We'll see much later this kind of quest and it's requirements are much better handled by a directed acyclic graph in Chapter \ref{chap-graphs}, but this example is fine for teaching lists.} first, and you must get the map to find the lair before you can physically travel there.

\subsection*{shopping list example}
While lists are defined by order, we don't necessarily  need to ascribe any meaning to the order.
Take a look at the shopping list below:




<Shopping List>
While bread is the first item on this list, being the first item in the shopping list in this case has no special meaning.  It's not the most important item on the list\footnote{obviously, that's the cookies}, nor is it necessarily the item I'm going to pick up first.

Where arrays and lists differ is that lists can grow to an arbitrary size, whereas

\begin{itemize}
	\item Lists can grow to any arbitrary size, but lists  
\end{itemize}

Data in a list is stored sequentially and array lists do this by creating a an 


\subsection{A note on terminology}
An \textbf{array list} is a type of list.  These are sometimes called dynamic arrays.

As mentioned in Section \ref{sec-python-arrays-lists}, Python doesn't have arrays.  If you've been programming in Python, you've been using an array list the entire time you've declared \texttt{[]}.  They are usually just called lists rather than array lists.


I will be using the Java nomenclature for the majority of the book as this allows me to be clear about the types and implementations of data structures.

\subsection{Lists in Java}

So what does

\subsubsection{An Aside about interfaces}
\subsubsection*{Here is the source code}



\section{Generics}

\subsection{What are they?}


Before we get to deep into lists, we need to have a discussion about generics. Generics are a way of restricting and specifying what types can go into a collection.

\subsection{But Why?}




\section{List Operations}


\subsection{Size} We need some easy way of knowing how big our lists are, if for no other reason than to make sure our  add and remove methods can figure our their valid indices.


\subsection{Add}  By default, we add items to the end of the list, but we can also add items to any index we want.

When we add an item at some specific index $ i $, the item at $ i $ and all indices to the right shift over one.  In other words, what was at $ i $ is now at $ i+1 $, what was at $ i+1 $ is at $ i+2 $,  and so on.

This also an understandable restriction to adding items to a list -  we cannot an item to any index greater than \texttt{myList.size() +1}.   Anything greater wouldn't be at the end of the list; it would be beyond it.  The same goes for negative indices.
 
<possible picture showing a legal an illegal add>

%We will cover this operation in more detail when we implement the add method for the arraylist 

\subsection{Remove}  We can remove Items from a list much in the same way we can add them. When removing an item at index I , all the indices to the right shift one over to the left 
\subsection{Get}
\subsection{Set}

\section{ArrayLists}
An array list, as you might have guessed, are lists built using \textit{arrays}.\footnote{Shockingly, many of the names we give things at this point actually make sense.}
They work by growing or shrinking the array\footnote{A lie.  As you'll see we don't actually change the size of an array;  we create a new array of the appropriate size and copy everything over} automatically as items are added or removed from the list, giving the illusion that the data structure can hold an arbitrary amount of data.

We'll go into the specifics of how this works in Section \ref{buildingArraylist}.


\subsubsection{Java's ArrayLists}
Java's arrayList
\subsubsection{Python's Lists}
Python's lists, such as below:
\begin{minted}{python3}
	l = [1,2,3] # this is a list, not an array!	
\end{minted}
are actually array lists! %TODO cite this https://docs.python.org/3/faq/design.html#how-are-lists-implemented-in-cpython

Python uses a different vocabulary for some of the methods we'll be implementing below.  
For example, take the action of adding an item to a list.
Python uses the \texttt{append} method to add an item to end of the list and \texttt{insert} to put an item into the middle of the list.
Java (who's vocabulary we'll be following), uses \texttt{add} for both these contexts. 









\section{Example Algorithms}


\begin{minted}{Java}
	public static <E> boolean isPermutation(List<E> listA, List<E> listB) {
		
		if(listA.size() != listB.size()) {
			return false;
		}
		for(int i  = 0; i < listA.size() ; i++){
			E item =  listA.get(i);
			int countA = 0;
			int countB = 0;
			
			for (E element : listA) {
				if(item.equals(element)){
					countA++;
				}
			}
			for (E element : listB) {
				if(item.equals(element)){
					countB++;
				}
			}
			if(countA != countB) {
				return false;
			}
		}
		return true;
	}
\end{minted}



\section{Building an ArrayList}
\label{buildingArraylist}
To truly understand how a data structure works we need to implement it ourselves.

\subsection{Caveats}

\subsubsection{MyArrayList.java}
We will not be implementing the \texttt{List} interface. We don't need to implement all the functions to get an understanding of how the fundamentals of an arraylist work.
Implementing the list interface would take up a hideous amount of physical space and get in the way of actually understanding the code.



\subsubsection{myArrayList.py}

For python, this will require some suspension of disbelief, as our array list will require using an array, and as previously discussed, arrays are shirked in favor of arraylists in python.  We'll be using a list and pretending it's an array. Silly?  Yes.  But it will keep our code compact and easier to understand.

%Remember, the standard python list is built in the C programming language.

\subsection{Instance Variables}

Believe it or not, we only need to keep track of three instance variables to get our arraylist working.

\begin{enumerate}
	\item[theData]  We need an array to actually store the items.  This is it.
	\item[size] Size here refers to the total number of items we have stored in the array.
	\item[capacity]  This is the number of items  the underlying \texttt{array} in our list can hold. It is the maximum size of the list before we have to do something about it. This is not strictly necessary as we can get it by querying \texttt{array}'s length. However, making it it's own variable will help with the readability.
\end{enumerate}

It is very easy to confuse size and capacity since they both deal with counting how many elements.  When I talk about \texttt{size}, I am talking about the number of items we have stored in the list we are making.  Capacity, on the other hand, depends on the length of the built-in array. 

\subsubsection{Java}


\begin{minted}{Java}
public class MyArrayList<E> {
	
	private E[] theData
	private int size;  // how many items are in the list
	private int capacity;  // how many items the underlying array can hold
	
	
\end{minted}

First, note the \texttt{<E>} after \texttt{MyArrayList}.  This means that we're saying:
\begin{itemize}
	\item MyArrayList is designed to hold a specific type of object.
	\item Every \texttt{E} we see is a placeholder for some type, which will be that same across the entire lifespan of the object.
\end{itemize}


\subsubsection{Python}
\begin{minted}{Python3}
class MyArrayList(object):
	pass
\end{minted}


\subsection{Constructor}

We need to set the variables to their initial values upon creating the arraylist.
The \texttt{size} will be 0, since we won't have any objects stored in it yet.
We will set the \texttt{initial} capacity to 10.  
It's a small number and thus won't create much wasted space if we don't fill up \texttt{theData}.
\texttt{theData} will be an empty array of \texttt{capacity} length.
If \texttt{theData} becomes full, we will create a bigger array to hold our items using the \texttt{reallocate()} method (Section \ref{
arraylist-reallocate})  %TODO Subection Depth


\subsubsection{Java}
With our constructor, we have one line of weird black magic in order to create an Array of \texttt{E[]}'s.
\begin{minted}{Java}
public MyArrayList(){
	size = 0;
	capacity = 10;
	theData = (E[]) new Object[10]; // this generates a warning
}
\end{minted}

So what's going on with the last line?
Typically, when creating an array, we would just say:

\begin{minted}{Java}
//doing this in the constructor gives us an error.
TYPE[] myArray  = new Type[desired_size];
\end{minted}

However, Java won't let you create new \texttt{E} objects since there's no telling what the constructors will be.  This rule extends to arrays of \texttt{E}, like so:
\begin{minted}{Java}
theData =  new E[10];
\end{minted}

However, when creating a new empty array of objects of any type, we're just making an array of nulls which will eventually be replaced by references to objects.  Thus, even though the Java compiler will yell at us about Type safety, we can instead create an array of \texttt{Object} and then tell , since all references to any types are the same size.

\begin{minted}{Java}
// creating one array of nulls and telling Java 
// its another type of array of nulls. 
theData = (E[]) new Object[10];
\end{minted}



Remember how Java and most modern programming languages deal with objects; if you're assigning an object to a variable, like in \texttt{Object o = new Object()}, we are storing a reference to that object.

Thus, when we add an item to a list, what really happens is we'll be adding a reference to it - the instructions on how to find it in memory.
\subsubsection{Python}
Python is fairly straightforward, with the caveat that we are pretending \texttt{theData} is an array, and not a list.

\begin{minted}{Python3}
class MyArrayList(object):
	def __init__(self):
		self.size = 0
		self.capacity = 10
		self.theData = [None]*self.capacity
\end{minted}

Since built-in lists in Python grow and shrink like we would expect a list to, we initialize \texttt{theData} with 10 \texttt{None} objects\footnote{This is the Python equivalent to the Java \texttt{null}.} to mimic the way an array would be initialized.


%\subsection{Placeholder toString}
%
%
%\subsubsection{Java}
%
%\subsubsection{Python}

\subsection{Size}
Now, we will add a size method to our list; fairly straightforward in \texttt{Java}.

\begin{minted}{Java}
public int size() {  
	return size;
}
\end{minted}

In Python, we can go ahead and use the built in \texttt{\_\_len\_\_} method.

\begin{minted}{Python3}
def __len__(self):
	return self.size
\end{minted}

Which can then be invoked with \texttt{len(myList)}.


It is worth reinforcing that by keeping track of the size in a variable, retrieving the size of our list is always $O(1)$

\subsection{Add}



In Python, these two \texttt{add} methods are called \texttt{append} and \texttt{insert} respectively, as python does not support method overloading.

\begin{minted}{Java}
public void add(int index, E item) {
	if(index < 0 || index > size) {
		throw new IndexOutOfBoundsException("Index " + index + " out of bounds.");
	}
	
	if(size == capacity) {  // O(n) time...sometimes.  Amortized over the cost of adding
		this.reallocate();
	}
	
	for(int i = size - 1; i >= index; i--) { //If adding to the end... constant
		E temp = theData[i];
		theData[i+1] = temp;
	}
	
	theData[index] = item;
	size++;
}
\end{minted}


\begin{minted}{Python3}
def insert(self, index, item):
	
	
	self.size += 1

\end{minted}





Finally, since adding to the end is an extremely common operation, we will overload our \texttt{add} method.
If our list is provided with only an \texttt{item}, as opposed to an \texttt{item} and an \texttt{index}, we will just add that \texttt{item} to the end.
Since we already wrote a perfectly good \texttt{add} method already that we know works, we'll just have our new method call that one.
\begin{minted}{Java}
public boolean add(E item) {
	this.add(size, item); // size is the last valid index
	return true; // What?
}
\end{minted}

Why are we returning \texttt{true} here?
The short answer is practice and consistency with future data structures.
The long answer is any \texttt{Collection} in Java has must have an \texttt{add} method and a \texttt{List} is type of \texttt{Collection}\footnote{Our \texttt{MyArrayList} isn't technically a \texttt{Collection} since we did not implement the \texttt{List} interface, but I digress.}.

\texttt{Collection} specifies that \texttt{add} must take in an \texttt{item} and return a \texttt{boolean}.
A \texttt{true} signals the \texttt{add} is successful.
A \texttt{false} signals that we could not add the \texttt{item}.
For example, this might happen with a \texttt{Set} (Chapter \ref{chap-sets})


On the other hand, our Adding at a specific index is unique to lists, and not part of collections,  and will always work. Therefore, there's no need to return a boolean. 

\subsubsection{Reallocation}
\label{arraylist-reallocate}
When we need to grow our arraylist
%TODO fix this code or ensure that capacity is doubled before copying
\begin{minted}{Java}


private void reallocate(){
	//doubles or 1.5x capacity
	//don't do +1 capacity
	E[] newData = (E[]) new Object[capacity];
	for(int i = 0; i < theData.length; i++) {
		newData[i] =  theData[i];
	}
	
	theData = newData;
	capacity = 2 * capacity;
	
}
\end{minted}

We want to double our capacity or at least increase it by 50\%, Rather than increasing it by a static number.
Consider if we increase the capacity by one each time we reallocated.  If we did that, we would have to reallocate every time we added a new item to the list.  This would mean that every time we add an item to list, add becomes a linear time - $O(n)$ - operation.

%TODO explain what happens to runtime amortization when we reallocate but double the size

\subsection{toString and \_str\_}
Now that we supposedly have a method for adding items into the list.


\subsection{Get and Set}
The \texttt{get} and \texttt{set} methods are fairly straightforward:
\begin{enumerate}
\item[\texttt{get} -] Given an \texttt{index}, retrieve the item stored at that \texttt{index}.
\item[\texttt{set} -] Given an \texttt{index}, replace the old item stored at that \texttt{index} with the provided \texttt{item}.
\end{enumerate}
%For \texttt{get}, given an \texttt{index}, retrieve the item stored at that \texttt{index}.
%For \texttt{set}, given an \texttt{index}, replace the old item stored at that \texttt{index} with the provided \texttt{item}.
\texttt{set} has one additional quirk, we also want to return the old item we're replacing, just in case the programmer wants to doing something with the old item.
This would obviate the need for pairing a \texttt{get} and \texttt{set} call with each other if we want to replace the old item, but do something else with it. 

For both \texttt{get} and \texttt{set}, we want to throw some kind of error if the provided index is out of bounds.
\subsubsection{Java}
Our \texttt{get} is fairly straightforward, but feel free to give more information with the error.
\begin{minted}{Java}
public E get(int index) {
	if(index  < 0 || index >= size) {
		throw new IndexOutOfBoundsException("Index " + index + " out of bounds.");
	}
	return theData[index];
}
\end{minted}

The same goes for our \texttt{set} method. 
\begin{minted}{Java}
public E set(int index, E item) {
	if(index < 0 || index >= size) {
		throw new IndexOutOfBoundsException("Index " + index + " out of bounds.");
	}
	E oldItem  = theData[index];
	theData[index] = item;
	
	return oldItem;
}
\end{minted}

\subsubsection{Python}

Python supports negative indices.  It's up to you whether or not you want to support that, but if you do you need to account for the \texttt{capacity} and the \texttt{size} of the internal list.


We can take advantage of some of the method calls built into python to make our \texttt{myarraylist} support indexing.


\subsection{Remove}




\section{Analysis}

\section{More Restrictive or Permissive Generics}

\section{A Few More Useful Methods}

\subsection{Constructors}
Java's \texttt{ArrayList} can optionally take in an integer as an argument

\subsection{Trim}

The \texttt{trim()} exists to mitigate the fact that arraylist capacity and size don't often match up.


On the other hand, Python will automatically optimize lists for you CITATION NEEDED.

For other languages, check to see if the equivalent exists

\subsection{Adding Multiple Items in One Invocation}
One common operation is to move or copy all the items from one list to another.
In Java, we can use the \texttt{addAll()} method, which takes any Java collection as a parameter and all the items in that collection to the object.

\begin{minted}{Java}
List<Integer> a =  new ArrayList<>();
List<Integer> b =  new ArrayList<>();
for(int i = 0; i <3; i++) { 
	a.add(i); 
}
for(int i = 3; i <6; i++) { 
	b.add(i); 
}
a.addAll(b);
System.out.println(a); // 0 to 5 inclusive
System.out.println(b); // [3, 4, 5]
\end{minted}

In Python, we can use the \texttt{extend()} method on anything that is iterable or use some clever slicing.  However, I would always recommend using the method call over the slice, since a method invocation is always more readable.


\begin{minted}{Python3}
a = [0, 1, 2]
b = [3, 4, 5]
c = a + b # creates a new list, which is not extend
a.extend(b) # adds all of b's items to a
a[len(a):] = b # does the same thing but unreadble.
\end{minted}
A common beginner mistake in Python is to try to extend a list by calling \texttt{append} on the list like so.

\begin{minted}{Python3}
a = [0, 1, 2]
b = [3, 4, 5]
a.append(b) # a is now [0, 1, 2, [3, 4, 5]]
\end{minted}



This adds the entire list a single item in the list.  